The non-random assembly of indirect interactions in  \nplant-pollinator networks

### 2022-07-25 Manuscript/Main_text/Main_text_Journal_of_Animal_Ecol.Rmd
Correction of the use of the term motif

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organized is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. We found a higher proportion of densely connected motifs, which could reflect that neutral effects are the predominant ecological process governing the structure of network motifs. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different floral visitor and plant groups on both trophic levels. For example, bees tend to appear in motif positions with a lower number of indirect interactions than the rest of floral visitor groups. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our results highlight the non-random structure of the key building blocks shaping plant-pollinators networks and the association of certain motifs with different groups of floral visitors and plants.

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions [i.e., the mediated effect between two species by a third species\; @holt1994], such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs [small sub-networks observed within a given network which are often referred as its building blocks\; @milo2002] allows to consider both direct and indirect interactions [@simmons2019]. The analysis of network motifs in plant-pollinator networks have revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

@@

Here, we used 60 empirical networks from 18 different studies and 14 countries to evaluate the structure of the meso-scale in plant-pollinator networks. From these networks we extracted plant and floral visitors species and determined the principal groups that define their main life-history strategies. To obtain plant groups, we used a comprehensive dataset that included floral, reproductive and vegetative traits compiled in a larger set of plant species. Floral visitor groups were divided by the main taxonomic groups that differed in life form and behaviour. Once we split the different plant-pollinator networks into their motif elements, we explored: (i) if there is a common invariant structural property in the overall motif networks (i.e., over- and under-represented motifs); (ii) which plant and floral visitor groups are over- and under-represented in different motif positions associated with specific ecological roles; and, (iii) if there are over- and under-represented plant and floral visitor group combinations within a motif.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into groups that summarise their main life-history strategies. After, we calculated the normalised motif frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different motif positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different motif positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among the patterns of connections of motif positions. Finally, we compared if there are differences in the observed frequencies of motif combinations from the empirical networks and the expected frequencies of motif combinations calculated as the product of the probabilities of the different plant and floral visitor groups within the motif.

@@

**Overall motif patterns**

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent motifs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. This approach was preferred since quantitative motif analysis have been developed just recently and more work is needed to fully understand and interpret the information that they can convey [@simmons2020]. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (Figs S2 and S3).

We calculated the frequency of all motifs up to five nodes (17 different motifs in total; see Fig. 1) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on motifs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network motifs. Throughout the entire manuscript, floral visitors occupy the top nodes of the different motifs and plants the bottom ones. To control for variation in network size, motif frequencies were normalised as a proportion of the total number of motifs within each motif class (i.e., 2-species motifs, 3-species motifs, 4-species motifs and 5-species motifs). In addition, we excluded 2-species (or nodes) motifs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different motifs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected motifs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

```{r, echo=FALSE, message=FALSE, cache=FALSE, fig.pos="t", warning=FALSE,out.width="75%",fig.align="center", fig.cap="\\textbf{Figure 1.} Adapted figure of Simmons et al., (2019). All possible motif network structures from two to five species in bipartite networks grouped by the different motif classes (2-species, 3-species, 4-species and 5-species). There is a total of 17 possible motifs with 46 different positions denoted within each node. The different motifs are aggregated by their average path length classification wich goes from densely connected motifs (strong motifs coloured in yellow) to less connected motifs (weak motifs coloured in red). Note that two node motifs are excluded from this classification.", fig.width=10,fig.height=8}
library(tidyverse)
library(RColorBrewer)
source("../../Scripts/Scripts_Alfonso/plot_motif_positionsV2.R")

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variability and provided highly similar or identical networks to the original ones. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controlling for the variation at the study level.

**Over- and under-represented groups on motif positions**

We calculated which plant and floral visitor groups were over- or under-represented in the different motif positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same motif size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with motif frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per motif position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average motif frequency per species group, after controlling for the variation at the study level. Finally, we visualized with the help of the _ComplexHeatmap_ package version _2.6.2_ [@gu2016] over- and under-representation of plant and floral visitor groups on the different motif positions.

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialisation (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2020] version 2.7.0.

**Over- and under-represented group combinations of motifs**

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same motifs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the motifs of the three networks with the highest number of links. To do so, for each of the 53,250 possible motif combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a motif $i$ (e.g., motif 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability ($p_i^E$) of a given motif combination $i$, we assumed that the probability of finding a given group $x$ in the position $\alpha$ is independent of the probabilities associated with other positions and groups in that motif. Therefore, the expected probability ($p_i^E$) of the motif combination $i$ is given by the product of the probabilities of its constituents, i.e., $p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$, where $p_i(x,\alpha)$ is the probability of finding a given group $x$ in the position $\alpha$ of $i$. Then, by controlling for the variation at the network level with an LMM (see below), we calculated the average number of times that the group $x$ appears in the position $\alpha$ (denoted as $n_i(x,\alpha)$). To obtain the expected probability $p_i(x,\alpha)$ we divided the number of times that the group $x$ appears in the position $\alpha$ by the the total number of times that any group was observed in that position (that is, $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$). The average number of times that the group $x$ appears in the position $\alpha$ ($n_i(x,\alpha)$) was calculated with a LMM for each position, where the response variable was the number of times that a group was observed at a given position in each real-world network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

Once we obtained $p_i^O$ and $p_i^E$, we used a simulation approach to determine whether the former is likely to come from the latter or not. This approach was preferred since the large number of possible motif combinations and the small probabilities for some of them advise against using an exact test of goodness-of-fit or a Chi-square one. Specifically, we created 1,000 random samples with repetition of possible motif combinations, where each sample contained 10 million elements and, for each combination, the probability of being selected was equal to its expected probability. From those random samples, we extracted the mean and the standard deviation of the expected probability of $i$, denoted as $E\left [ p_i^E \right ]$ and $\sigma\left [ p_i^E \right ]$, respectively, and calculated the z-scores of $p_i^O$ as $z_i^O = \left ( p_i^O - E\left [ p_i^E \right ] \right )/\sigma \left [ p_i^E \right ]$, for those motif combinations with $p_i^O > 0$. According to the usual interpretation of z-scores, combinations with $z_i^O > 1.96$ are over-represented, whereas those with $z_i^O < -1.96$ are under-represented, at the 95$\%$  confidence level. Notice that we focused on combinations with $p_i^O > 10^{-7}$ because we do not have enough numerical resolution to accurately detect whether combinations that took place only once are under-represented (due to the limited size of our random samples).

@@

**Overall motif patterns**

Most motifs were under- or over-represented (close to the 1st and 99th percentile, respectively) in the comparison between empirical and simulated networks (Fig. 2). Motifs 3, 5, 9, 10 and 14 were under-represented in empirical networks, that is, all were close to the 1st percentile and under the 25th percentile. Interestingly, four out of five of these motifs belonged to the largest path length classification (i.e., core-peripheral). In addition, motifs 2, 6, 7, 16 and 17 were over-represented, all over the 75th percentile. In contrast to the under-represented motifs, over-represented motifs belonged to the two shortest path length groups (i.e., complete and fan). The remaining motifs (i.e., 4, 8, 11, 12, 13, and 15) were between the 25th and 75th percentile.

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.width=4.5,fig.height=6.5, fig.cap="**Figure 2.** Comparison of motif frequencies between empirical and simulated networks grouped by average path length (plots a, b, c and d) as determined in Simmons et al. (2020). This is shown with the mean percentage of motif frequencies in empirical networks that were over the motif frequencies of the simulated ones (percentiles). This was done by network (light blue transparent dots) and then averaged for all networks (black dots with error bars that correspond to the standard deviation)."}

@@

**Over- and under-represented groups on motif positions**

@@

The comparison of the plant and floral visitor group frequencies per motif position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (<25th) groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented groups, respectively. Notably, the differences across groups were more marked for floral visitors than for plant groups (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors groups on the different motif positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented groups than others, thus from most over- to under-represented groups on the different motif positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant groups we found over- and under-representation on different motif positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on motif positions that involved specialization (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised motif positions with low number of indirect interactions (Fig. S6 A). Although plants did not show differences with null models on their frequencies, all plant but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised motif positions (Fig. S6 B). Second, regarding the number of indirect interactions, bees were over-represented in motif positions with low number of indirect interactions (Fig. S7 A) and the rest of floral visitor groups showed no differences across positions in comparison with simulated networks. For plants, all groups showed a slight trend to be over-represented on motif positions with higher number of indirect interactions (Fig. S7 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on motif positions associated with low number interactions in comparison with the other floral visitor and plant groups (Figs S8 and S9).

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.height=8,fig.width=8,fig.align = 'center', out.width="80%",fig.cap="\\textbf{Figure 3.} Heatmap indicating under- and over-representation of floral visitor and plant groups in the different motif positions (numbers on the right of each heatmap). The different motif positions are dividied by the average path length clasification determined by Simmons et al. (2020) and they are ordered from the most densely connected motifs (strong or complete) to the less connected ones (weak or core-peripheral). The superior dendrogram indicates the differences across groups with the more separated groups showing larger differences."}

@@

**Over- and under-represented group combinations of motifs**

>>>>

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of animal-plant mutualistic networks. Detecting how direct and indirect interactions between animals and plants are organized is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the meso-scale, the intermediate level of network complexity between the species and the full network, can capture this important information. The meso-scale describes network subgraphs representing patterns of direct and indirect interactions between a small number of species and when these network subgraphs differ statistically from a benchmark, they are often referred to as "network motifs". Despite motifs can capture relevant ecological information of species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some network subgraphs are consistently under- or over-represented, suggesting the presence of worldwide network motifs in plant-pollinator networks. In addition, we found a higher proportion of densely connected network subgraphs, which could reflect that neutral effects are the predominant ecological process governing the structure of the meso-scale. We found that distinct subgraph positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different floral visitor and plant groups on both trophic levels. For example, bees tend to appear in subgraph positions with a lower number of indirect interactions than the rest of floral visitor groups. Interestingly, the realized combinations of different species groups within a network subgraph can not be retrieved from their joint probability distributions, indicating that plant and floral visitor group combinations on network subgraphs are not random. Our results highlight the non-random structure of the key building blocks shaping plant-pollinators networks and the association of certain network subgraphs and their positions with specific groups of floral visitors and plants.

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions [i.e., the mediated effect between two species by a third species\; @holt1994], such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the analysis of network subgraphs at the meso-scale, the intermediate level of complexity between the full network and the species level, allows to consider both direct and indirect interactions [@simmons2019]. The network subgraphs that significantly depart from an expected benchmark are often defined as 'network motifs' [@milo2002]. The analysis of network motifs in plant-pollinator networks have revealed that the different drivers that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented network subgraphs).

@@

Here, we used 60 empirical networks from 18 different studies and 14 countries to evaluate the structure of the meso-scale in plant-pollinator networks. From these networks we extracted plant and floral visitors species and determined the principal groups that define their main life-history strategies. To obtain plant groups, we used a comprehensive dataset that included floral, reproductive and vegetative traits compiled in a larger set of plant species. Floral visitor groups were divided by the main taxonomic groups that differed in life form and behaviour. Once we split the different plant-pollinator networks into their subgraphs elements, we explored: (i) if there is a common invariant structural property in the overall subgraphs from these networks (i.e., over- and under-represented subgraphs); (ii) which plant and floral visitor groups are over- and under-represented in different subgraphs positions associated with specific ecological roles; and, (iii) if there are over- and under-represented plant and floral visitor group combinations within a subgraph.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into groups that summarise their main life-history strategies. After, we calculated the normalised subgraph frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different subgraph positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different subgraph positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among the patterns of connections of subgraph positions. Finally, we compared if there are differences in the observed frequencies of subgraph combinations from empirical networks and the expected frequencies calculated as the product of the probabilities of the different plant and floral visitor groups within the network subgraph.

@@

**Overall network subgraphs patterns**

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent network subgraphs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. This approach was preferred since quantitative subgraph analysis have been developed just recently and more work is needed to fully understand and interpret the information that they can convey [@simmons2020]. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (Figs S2 and S3).

We calculated the frequency of all subgraphs up to five nodes (17 different subgraphs in total; see Fig. 1) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on subgraphs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network structures. Throughout the entire manuscript, floral visitors occupy the top nodes of the different network subgraphs and plants the bottom ones. To control for variation in network size, subgraph frequencies were normalised as a proportion of the total number of subgraphs within each subgraph class (i.e., 2-species, 3-species, 4-species and 5-species network subgraphs). In addition, we excluded 2-species (or nodes) subgraphs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different subgraphs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected network subgraphs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

```{r, echo=FALSE, message=FALSE, cache=FALSE, fig.pos="t", warning=FALSE,out.width="75%",fig.align="center", fig.cap="\\textbf{Figure 1.} Adapted figure of Simmons et al., (2019). All possible network subgraphs from two to five species in bipartite networks grouped by the different subgraph classes (2-species, 3-species, 4-species and 5-species). There is a total of 17 possible subgraphs with 46 different positions denoted within each node. The different subgraphs are aggregated by their average path length classification wich goes from densely connected (strong subgraphs coloured in yellow) to less connected network structures (weak subgraphs coloured in red). Note that two node subgraphs are excluded from this classification.", fig.width=10,fig.height=8}
library(tidyverse)
library(RColorBrewer)
source("../../Scripts/Scripts_Alfonso/plot_motif_positionsV2.R")

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variability and provided highly similar or identical networks to the original ones. After extracting the subgraph frequencies from the simulated networks, for each subgraph type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed subgraph frequencies. Subgraphs whose percentile is close to 0 or 100 are under- or over-represented in empirical networks, respectively, and indicate the presence of network motifs as they differ statistically from the network structures generated with simulated networks. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per subgraph with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed subgraph percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average subgraph frequency, but controlling for the variation at the study level.

**Over- and under-represented groups on subgraph positions**

We calculated which plant and floral visitor groups were over- or under-represented in the different subgraph positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same subgraph size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with subgraph frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per subgraph position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average subgraph frequency per species group, after controlling for the variation at the study level. Finally, we visualized with the help of the _ComplexHeatmap_ package version _2.6.2_ [@gu2016] over- and under-representation of plant and floral visitor groups on the different subgraph positions.

To understand if the different plant and floral visitor groups are associated with subgraph positions that represent different network roles, we characterized the roles of each subgraph position for plant and floral visitor groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each subgraph position following @poisot2015 as $s_i = (L - l_i) / (L - 1)$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given subgraph. Values of the vector $si$, range between 1, complete specialisation (1 partner within the subgraph) and 0 complete generalism (all possible partners within the subgraph). Then, because all nodes within a subgraph are connected, we calculated the number of indirect interactions ($n_i$) for each subgraph position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given subgraph. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and subgraph position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor subgraph position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2020] version 2.7.0.

**Over- and under-represented group combinations of subgraphs**

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same subgraphs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the subgraphs of the three networks with the highest number of links. To do so, for each of the 53,250 possible subgraph combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a subgraph $i$ (e.g., subgraph 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability ($p_i^E$) of a given subgraph combination $i$, we assumed that the probability of finding a given group $x$ in the position $\alpha$ is independent of the probabilities associated with other positions and groups in that subgraph. Therefore, the expected probability ($p_i^E$) of the subgraph combination $i$ is given by the product of the probabilities of its constituents, i.e., $p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$, where $p_i(x,\alpha)$ is the probability of finding a given group $x$ in the position $\alpha$ of $i$. Then, by controlling for the variation at the network level with an LMM (see below), we calculated the average number of times that the group $x$ appears in the position $\alpha$ (denoted as $n_i(x,\alpha)$). To obtain the expected probability $p_i(x,\alpha)$ we divided the number of times that the group $x$ appears in the position $\alpha$ by the the total number of times that any group was observed in that position (that is, $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$). The average number of times that the group $x$ appears in the position $\alpha$ ($n_i(x,\alpha)$) was calculated with a LMM for each position, where the response variable was the number of times that a group was observed at a given position in each real-world network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

Once we obtained $p_i^O$ and $p_i^E$, we used a simulation approach to determine whether the former is likely to come from the latter or not. This approach was preferred since the large number of possible subgraph combinations and the small probabilities for some of them advise against using an exact test of goodness-of-fit or a Chi-square one. Specifically, we created 1,000 random samples with repetition of possible subgraph combinations, where each sample contained 10 million elements and, for each combination, the probability of being selected was equal to its expected probability. From those random samples, we extracted the mean and the standard deviation of the expected probability of $i$, denoted as $E\left [ p_i^E \right ]$ and $\sigma\left [ p_i^E \right ]$, respectively, and calculated the z-scores of $p_i^O$ as $z_i^O = \left ( p_i^O - E\left [ p_i^E \right ] \right )/\sigma \left [ p_i^E \right ]$, for those motif combinations with $p_i^O > 0$. According to the usual interpretation of z-scores, combinations with $z_i^O > 1.96$ are over-represented, whereas those with $z_i^O < -1.96$ are under-represented, at the 95$\%$  confidence level. Notice that we focused on combinations with $p_i^O > 10^{-7}$ because we do not have enough numerical resolution to accurately detect whether combinations that took place only once are under-represented (due to the limited size of our random samples).

@@

**Overall network subgraphs patterns**

Most network subgraphs were under- or over-represented (close to the 1st and 99th percentile, respectively) in the comparison between empirical and simulated networks (Fig. 2). Network subgraphs 3, 5, 9, 10 and 14 were under-represented in empirical networks, that is, all were close to the 1st percentile and under the 25th percentile. Interestingly, four out of five of these motifs belonged to the largest path length classification (i.e., core-peripheral). In addition, network subgraphs 2, 6, 7, 16 and 17 were over-represented, all over the 75th percentile. In contrast to the under-represented motifs, over-represented motifs belonged to the two shortest path length groups (i.e., complete and fan). The remaining network subgraphs (i.e., 4, 8, 11, 12, 13, and 15) were between the 25th and 75th percentile.

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.width=4.5,fig.height=6.5, fig.cap="**Figure 2.** Comparison of network subgraph frequencies between empirical and simulated networks grouped by average path length (plots a, b, c and d) as determined in Simmons et al. (2020). This is shown with the mean percentage of subgraph frequencies in empirical networks that were over the subgraph frequencies of the simulated ones (percentiles). This was done by network (light blue transparent dots) and then averaged for all networks (black dots with error bars that correspond to the standard deviation)."}

@@

**Over- and under-represented groups on subgraph positions**

@@

The comparison of the plant and floral visitor group frequencies per subgraph position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (<25th) groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented groups, respectively. Notably, the differences across groups were more marked for floral visitors than for plant groups (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors groups on the different subgraph positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented groups than others, thus from most over- to under-represented groups on the different subgraph positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant groups we found over- and under-representation on different subgraph positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on subgraph positions that involved specialization (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised subgraph positions with low number of indirect interactions (Fig. S6 A). Although plants did not show differences with null models on their frequencies, all plant but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised subgraph positions (Fig. S6 B). Second, regarding the number of indirect interactions, bees were over-represented in subgraph positions with low number of indirect interactions (Fig. S7 A) and the rest of floral visitor groups showed no differences across positions in comparison with simulated networks. For plants, all groups showed a slight trend to be over-represented on subgraph positions with higher number of indirect interactions (Fig. S7 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on subgraph positions associated with low number interactions in comparison with the other floral visitor and plant groups (Figs S8 and S9).

```{r, echo=FALSE, message=FALSE, cache=FALSE,  warning=FALSE, fig.height=8,fig.width=8,fig.align = 'center', out.width="80%",fig.cap="\\textbf{Figure 3.} Heatmap indicating under- and over-representation of floral visitor and plant groups in the different subgraph positions (numbers on the right of each heatmap). The different subgraph positions are dividied by the average path length clasification determined by Simmons et al. (2020) and they are ordered from the most densely connected network structures (strong or complete) to the less connected ones (weak or core-peripheral). The superior dendrogram indicates the differences across groups with the more separated groups showing larger differences."}

@@

**Over- and under-represented group combinations of subgraphs**
